Bayesian Sparsification of $\cplx$-valued networks

### 2019-11-07 paper/paper.tex
cplx-gaussian overhaul

@@

A random vector $z\in \cplx^m$ is said to have complex Gaussian distribution, if its
characteristic function is
$$
z \sim \mathcal{N}^{\cplx}_m(z \vert \mu, \Gamma, C)
  \Leftrightarrow
  \mathbb{E}_{z} e^{j \Re{(t^{\hop} z)}}
  = \exp{\{
    j \Re{(t^{\hop} \mu)}
    - \tfrac14 \bigl(
      t^{\hop} \Gamma t + \Re{(t^{\hop} C \conj{t})}
    \bigr)
  \}}
  \,, $$
where $\mu \in \cplx^m$ is the mean, $\Gamma, C \in \cplx^{m\times m}$ are complex covariance
and relation matrices satisfying $C^\top = C$, $\Gamma^{\hop} = \Gamma$, $\Gamma \succeq 0$,
and $
  \conj{\Gamma} \succeq C^{\hop} \Gamma^{-1} C
$. The matrices $\Gamma$ and $C$ correspond to $
  \mathbb{E} (z - \mu)(z - \mu)^{\hop}
$ and $
  \mathbb{E} (z - \mu)(z - \mu)^\top
$, respectively. Here ${\cdot}^{\hop}$ denotes Hermitian conjugate, i.e. $
  A^{\hop} = (\conj{A})^\top
$, ${\cdot}^{\top}$ -- the matrix transpose, and $\conj{\cdot}$ -- the complex conjugation
elementwise.
%
% The intuition behind them is more apparent in the complex univariate case ($m=1$):
% here $\Gamma = \sigma^2 \geq 0$, and $C$ can be represented as $\sigma^2 \rho \in \cplx$ with $
%   \lvert \rho \rvert \leq 1
% $. Therefore $\sigma^2$ affects the overall ``concentration'' of the complex random variable,
% whereas $\rho$ is responsible for the strength and ``direction'' of the linear relation between
% the real and imaginary parts, \cite{lapidoth_capacity_2003}.
% $$
% \frac{\sigma^2}2
%   \begin{pmatrix}
%     1 + \Re{\rho} & \Im{\rho} \\
%     \Im{\rho} & 1 - \Re{\rho}
%   \end{pmatrix}
%   = \frac{\sigma^2}2 I
%   + \frac{\sigma^2 \lvert \rho \rvert}2 
%   \begin{pmatrix}
%     \cos\theta & \sin\theta \\
%     \sin\theta & - \cos\theta
%   \end{pmatrix}
% $$

A characterization of a $\cplx$-Gaussian vector, more useful for the current study, is via the
equivalence with the paired real multivariate Gaussian distribution:
\begin{multline}  \label{eq:cn-paired-real-density}
  z \sim \mathcal{N}^{\cplx}_m(\mu, \Gamma, C)
    \\\Leftrightarrow
    \begin{pmatrix}
      \Re z \\ \Im z
    \end{pmatrix}
      \sim \mathcal{N}_{2 m}\biggl(
        \begin{pmatrix}
          \Re \mu \\ \Im \mu
        \end{pmatrix},
        \frac12
        \begin{pmatrix}
          \Re{(\Gamma + C)} & - \Im{(\Gamma - C)} \\
          \Im{(\Gamma + C)} &   \Re{(\Gamma - C)}
        \end{pmatrix}
      \biggr)
    \,.
\end{multline}
It is possible to recover the $\cplx$ covariance and relation matrices from a block covariance
matrix $\Sigma$ of a \textit{non-degenerate} paired real Gaussian vector through
$$
\begin{pmatrix}
  \Sigma_{rr} & \Sigma_{ri} \\ 
  \Sigma_{ir} & \Sigma_{ii} 
\end{pmatrix}
  = \frac12
    \begin{pmatrix}
      \Re{(\Gamma + C)} & - \Im{(\Gamma - C)} \\
      \Im{(\Gamma + C)} &   \Re{(\Gamma - C)}
    \end{pmatrix}
  \, \Leftrightarrow
  \,
  \begin{cases}
    \Gamma
      = \Sigma_{rr} + \Sigma_{ii} + j (\Sigma_{ir} - \Sigma_{ri})
    \,, \\
    C = \Sigma_{rr} - \Sigma_{ii} + j (\Sigma_{ir} + \Sigma_{ri}) 
  \end{cases}
  $$

A $\cplx$-Gaussian random vector $z$ is called \textit{proper}, when $C = 0$, i.e. the random
vector $z$ and its conjugate $\conj{z}$ are uncorrelated. This implies that in the paired real
representation $
  \Sigma_{ir} = -\Sigma_{ri} = -\Sigma_{ir}^\top
$ and $\Sigma_{rr} = \Sigma_{ii}$, i.e. real and imaginary component of the same random element
are uncorrelated, since skew-symmetric matrices necessarily have zero-diagonal.

The $\cplx$-Gaussian family of distributions is closed under affine transformations:
if $
  x \sim \mathcal{N}_m^{\cplx}(\mu, \Gamma, C)
$ and $y = A x + b$, for $
  A \in \cplx^{n\times m}
$ and $b \in \cplx^{n}$, then $
  y \sim \mathcal{N}_n^{\cplx}(A\mu + b, A \Gamma A^\hop, A C A^\top)
$. This follows from the effects of the affine transformation on the characteristic
function.

@@

The expression of Kullback-Leibler divergence of $q(\omega)$ from an arbitrary prior $\pi(\omega)$
necessarily involves the (differential) entropy of $q$:
\begin{equation*}
  KL\bigl(
    q(\omega) \| \pi(\omega)
  \bigr)
    % = \mathbb{E}_{\omega \sim q} \log{q(\omega)} - \mathbb{E}_{\omega \sim q} \log{\pi(\omega)}
    = - \mathbb{H}(q) - \mathbb{E}_{\omega \sim q} \log{\pi(\omega)}
    \,.
\end{equation*}
The entropy of a real multivariate Gaussian $
  x \sim q(x) = \mathcal{N}_n(x \vert \mu, \Sigma)
$ is
$$
\mathbb{H}(q)
  = - \mathbb{E}_{x\sim q(x)} \log{q(x)}
  % = \tfrac12 \log \det{(2 \pi \Sigma)}
  % + \tfrac12 \mathop{tr}{(
  %     \Sigma^{-1}
  %     \mathbb{E}_{x\sim q(x)}
  %       (x - \mu) (x - \mu)^\top
  %   )}
  = \tfrac12 \log \det{(2 \pi e \Sigma)}
  \,, $$
whence follows the expression for the entropy of a general $\cplx$-Gaussian random vector
$
  z \sim q(z)
    = \mathcal{N}^{\cplx}_n(z\vert \mu, \Gamma, C)
$
\begin{align*} \label{eq:c-gauss-entropy-derivation}
  \mathbb{H}(q)
    &
    % = \tfrac12 \log \det{(2 \pi e \Sigma)}
    = \tfrac12 \log \det{\biggl(
      2 \pi e \tfrac12
      \begin{pmatrix}
        \Re{(\Gamma + C)} & - \Im{(\Gamma - C)} \\
        \Im{(\Gamma + C)} &   \Re{(\Gamma - C)}
      \end{pmatrix}
    \biggr)}
    \notag \\
    % &
    % = \begin{bmatrix}
    %   r_1 + j r_2 \to r_1 \\  % sufficient for C=0 case
    %   j r_1 + r_2 \to r_2     % j conj of transformation for r1, det x 2
    % \end{bmatrix}
    % = \tfrac12 \log \det{\biggl(
    %   % \det{[I & j I \\ j I & I]} = \det{I} \det{I - jI I^{-1} jI} = \det{2 I}
    %   % left apply : thus dividing by \sqrt{2}
    %   \tfrac1{\sqrt{2}} 2 \pi e \tfrac12
    %   \begin{pmatrix}
    %     \Gamma + C & j (\Gamma - C) \\
    %     j (\conj{\Gamma} + \conj{C}) & \conj{\Gamma} - \conj{C}
    %   \end{pmatrix}
    % \biggr)}
      % % Implicitly using the fact that for a non-singular $\Omega$ we have $
      % %   \det{A}
      % %     = \tfrac{\det{(\Omega A)}}{\det{\Omega}}
      % %     = \tfrac{\det{(A \Omega)}}{\det{\Omega}}
      % % $.
    % \notag \\
    % &
    % = \begin{bmatrix}
    %   c_1 - j c_2 \to c_1 \\  % -j conj of transformation for c2
    %   c_2 - j c_1 \to c_2     % sufficient for C=0 case
    % \end{bmatrix}
    % = \tfrac12 \log \det{\biggl(
    %   % \det{[I & - j I \\ - j I & I]} = \det{I} \det{I - (-j)I I^{-1} (-j)I} = \det{2 I}
    %   % right apply : thus dividing by \sqrt{2}
    %   \tfrac12 2 \pi e \tfrac12
    %   \begin{pmatrix}
    %     2 \Gamma     & - 2 j C \\
    %     2 j \conj{C} & 2 \conj{\Gamma}
    %   \end{pmatrix}
    % \biggr)}
    % \notag \\
    % &
    % = \begin{bmatrix}
    %   r_2 - j \conj{C} \Gamma^{-1} r_1 \to r_2
    % \end{bmatrix}
    % = \tfrac12 \log \det{\biggl(
    %   % \det{[I & 0 \\ - j \conj{C} \Gamma^{-1} & I]} = \det{I} = 1
    %   \pi e
    %   \begin{pmatrix}
    %     \Gamma & - j C \\
    %     0 & \conj{\Gamma} - \conj{C} \Gamma^{-1} C
    %   \end{pmatrix}
    % \biggr)}
    % \notag \\
    &
    = \tfrac12 \log \det{(\pi e \Gamma)} \det{(\pi e (\conj{\Gamma} - \conj{C} \Gamma^{-1} C))}
    \,.
\end{align*}
The entropy of a proper $\cplx$-Gaussian random vector ($C = 0$) is therefore 
\begin{equation}  \label{eq:cn-proper-entropy}
  % determinant commutes with complex conjugation, since it is multilinear
  \mathbb{H}(q)
    % = \tfrac12 \log \det{(\pi e \Gamma)} \det{(\pi e (\conj{\Gamma} - \conj{C} \Gamma^{-1} C))}
    = \tfrac12 \log \det{(\pi e \Gamma)} \det{(\pi e \conj{\Gamma})}
    % = \tfrac12 \log \det{(\pi e \Gamma)} \conj{\det{(\pi e \Gamma)}}
  %  = \tfrac12 \log \bigl\lvert \det{(\pi e \Gamma)} \bigr\rvert^2
    = \log \bigl\lvert \det{(\pi e \Gamma)} \bigr\rvert
    \,.
\end{equation}

>>>>

%notation
We first clarify some notation: $M^{\top}$ is the matrix transpose, $\conj{M}$ is elementwise
complex conjugation, and $M^{\hop} = (\conj{M})^\top$ denotes Hermitian conjugate.

@@

A random vector $z\in \cplx^m$ has complex Gaussian distribution, $
  z \sim \mathcal{N}^{\cplx}_m(\mu, \Gamma, C)
$, with mean $\mu \in \cplx^m$ and $\cplx^{m\times m}$ covariance and relation
matrices $\Gamma$ and $C$ are, respectively, if 
\begin{equation}  \label{eq:cn-paired-real-density}
  \begin{pmatrix}
    \Re z \\ \Im z
  \end{pmatrix}
    \sim \mathcal{N}_{2 m}\biggl(
      \begin{pmatrix}
        \Re \mu \\ \Im \mu
      \end{pmatrix},
      \frac12
      \begin{pmatrix}
        \Re{(\Gamma + C)} & - \Im{(\Gamma - C)} \\
        \Im{(\Gamma + C)} &   \Re{(\Gamma - C)}
      \end{pmatrix}
    \biggr)
    \,,
\end{equation}
provided they satisfy $\Gamma \succeq 0$, $\Gamma^{\hop} = \Gamma$, $C^\top = C$, and $
  \conj{\Gamma} \succeq C^{\hop} \Gamma^{-1} C
$. This distribution possesses the usual properties of $\real$-valued Gaussians, including
if $y = A z + b$ for $
  A \in \cplx^{n \times m}
$ and $b \in \cplx^{n}$, then $
  y \sim \mathcal{N}_n^{\cplx}\bigl(
      A\mu + b, A \Gamma A^{\hop}, A C A^\top
    \bigr)
$. The entropy of a $\cplx$-Gaussian distribution $
  q = \mathcal{N}^{\cplx}_m(\mu, \Gamma, C)
$ is
\begin{align} \label{eq:c-gauss-entropy-derivation}
  \mathbb{H}(q)
    & = - \mathbb{E}_{z \sim q} \log{q(z)}
    \notag \\
    &
    % entropy of \mathcal{N}_n(\mu, \Sigma)
    %   % = \tfrac12 \log \det{(2 \pi \Sigma)}
    %   % + \tfrac12 \mathop{tr}{(
    %   %     \Sigma^{-1}
    %   %     \mathbb{E}_{x\sim q(x)}
    %   %       (x - \mu) (x - \mu)^\top
    %   %   )}
    %   = \tfrac12 \log \det{(2 \pi e \Sigma)}
    = \tfrac12 \log \det{\biggl(
      2 \pi e \tfrac12
      \begin{pmatrix}
        \Re{(\Gamma + C)} & - \Im{(\Gamma - C)} \\
        \Im{(\Gamma + C)} &   \Re{(\Gamma - C)}
      \end{pmatrix}
    \biggr)}
    \notag \\
    % &
    % = \begin{bmatrix}
    %   r_1 + j r_2 \to r_1 \\  % sufficient for C=0 case
    %   j r_1 + r_2 \to r_2     % j conj of transformation for r1, det x 2
    % \end{bmatrix}
    % = \tfrac12 \log \det{\biggl(
    %   % \det{[I & j I \\ j I & I]} = \det{I} \det{I - jI I^{-1} jI} = \det{2 I}
    %   % left apply : thus dividing by \sqrt{2}
    %   \tfrac1{\sqrt{2}} 2 \pi e \tfrac12
    %   \begin{pmatrix}
    %     \Gamma + C & j (\Gamma - C) \\
    %     j (\conj{\Gamma} + \conj{C}) & \conj{\Gamma} - \conj{C}
    %   \end{pmatrix}
    % \biggr)}
      % % Implicitly using the fact that for a non-singular $\Omega$ we have $
      % %   \det{A}
      % %     = \tfrac{\det{(\Omega A)}}{\det{\Omega}}
      % %     = \tfrac{\det{(A \Omega)}}{\det{\Omega}}
      % % $.
    % \notag \\
    % &
    % = \begin{bmatrix}
    %   c_1 - j c_2 \to c_1 \\  % -j conj of transformation for c2
    %   c_2 - j c_1 \to c_2     % sufficient for C=0 case
    % \end{bmatrix}
    % = \tfrac12 \log \det{\biggl(
    %   % \det{[I & - j I \\ - j I & I]} = \det{I} \det{I - (-j)I I^{-1} (-j)I} = \det{2 I}
    %   % right apply : thus dividing by \sqrt{2}
    %   \tfrac12 2 \pi e \tfrac12
    %   \begin{pmatrix}
    %     2 \Gamma     & - 2 j C \\
    %     2 j \conj{C} & 2 \conj{\Gamma}
    %   \end{pmatrix}
    % \biggr)}
    % \notag \\
    % &
    % = \begin{bmatrix}
    %   r_2 - j \conj{C} \Gamma^{-1} r_1 \to r_2
    % \end{bmatrix}
    % = \tfrac12 \log \det{\biggl(
    %   % \det{[I & 0 \\ - j \conj{C} \Gamma^{-1} & I]} = \det{I} = 1
    %   \pi e
    %   \begin{pmatrix}
    %     \Gamma & - j C \\
    %     0 & \conj{\Gamma} - \conj{C} \Gamma^{-1} C
    %   \end{pmatrix}
    % \biggr)}
    % \notag \\
    &
    = \tfrac12 \log \det{(\pi e \Gamma)} \det{(\pi e (\conj{\Gamma} - \conj{C} \Gamma^{-1} C))}
    \,.
\end{align}
A $\cplx$-Gaussian random vector $z$ is \textit{proper}, if $z$ and $\conj{z}$ are uncorrelated,
i.e. $C = 0$, since the matrices $\Gamma$ and $C$ correspond to $
  \mathbb{E} (z - \mu)(z - \mu)^{\hop}
$ and $
  \mathbb{E} (z - \mu)(z - \mu)^{\top}
$, respectively. From \eqref{eq:c-gauss-entropy-derivation} its entropy is therefore 
\begin{equation}  \label{eq:cn-proper-entropy}
  % determinant commutes with complex conjugation, since it is multilinear
  \mathbb{H}(q)
    % = \tfrac12 \log \det{(\pi e \Gamma)} \det{(\pi e (\conj{\Gamma} - \conj{C} \Gamma^{-1} C))}
    = \tfrac12 \log \det{(\pi e \Gamma)} \det{(\pi e \conj{\Gamma})}
    % = \tfrac12 \log \det{(\pi e \Gamma)} \conj{\det{(\pi e \Gamma)}}
  %  = \tfrac12 \log \bigl\lvert \det{(\pi e \Gamma)} \bigr\rvert^2
    = \log \bigl\lvert \det{(\pi e \Gamma)} \bigr\rvert
    \,.
\end{equation}

the characteristic function is
$$
z \sim \mathcal{N}^{\cplx}_m(z \vert \mu, \Gamma, C)
  \Leftrightarrow
  \mathbb{E}_{z} e^{j \Re{(t^{\hop} z)}}
  = \exp{\{
    j \Re{(t^{\hop} \mu)}
    - \tfrac14 \bigl(
      t^{\hop} \Gamma t + \Re{(t^{\hop} C \conj{t})}
    \bigr)
  \}}
  \,, $$

Not every $\real^{2m}$ Gaussian corresponds to a $\cplx^m$ Gaussian vecor,
but when it does the covariance and relation can be recovered through
$$
\Sigma
  = \begin{pmatrix}
    \Sigma_{rr} & \Sigma_{ri} \\ 
    \Sigma_{ir} & \Sigma_{ii} 
  \end{pmatrix}
  \, \Leftrightarrow
  \,
  \begin{cases}
    \Gamma
      = \Sigma_{rr} + \Sigma_{ii} + j (\Sigma_{ir} - \Sigma_{ri})
    \,, \\
    C = \Sigma_{rr} - \Sigma_{ii} + j (\Sigma_{ir} + \Sigma_{ri}) 
  \end{cases}
  $$

This implies that $
  \Sigma_{ir}
  % = -\Sigma_{ri}
  = -\Sigma_{ir}^\top
$ and $\Sigma_{rr} = \Sigma_{ii}$, i.e. real and imaginary component of the same random element
are uncorrelated, since skew-symmetric matrices necessarily have zero-diagonal.

If $M \in \real^{m\times m}$, $M \succeq 0$ and $M^\top = M$, then
$$
  (u - j v)^\top M (u + j v)
    = u^\top M u + v^\top M v + j (u^\top M v - v^\top M u)
    % = u^\top M u + v^\top M v + j (u^\top M v - u^\top M^\top v)
    = u^\top M u + v^\top M v \geq 0
\,. $$
if $M^\top = - M$, then
$$
  u^\top M v - v^\top M u - j (u^\top M u + v^\top M v)
  % = u^\top M v - u^\top M^\top v - j (u^\top M u + v^\top M^\top v)
  = u^\top M v + u^\top M v - j (u^\top M u - v^\top M v)
  \,. $$

%
% The intuition behind them is more apparent in the complex univariate case ($m=1$):
% here $\Gamma = \sigma^2 \geq 0$, and $C$ can be represented as $\sigma^2 \rho \in \cplx$ with $
%   \lvert \rho \rvert \leq 1
% $. Therefore $\sigma^2$ affects the overall ``concentration'' of the complex random variable,
% whereas $\rho$ is responsible for the strength and ``direction'' of the linear relation between
% the real and imaginary parts, \cite{lapidoth_capacity_2003}.
% $$
% \frac{\sigma^2}2
%   \begin{pmatrix}
%     1 + \Re{\rho} & \Im{\rho} \\
%     \Im{\rho} & 1 - \Re{\rho}
%   \end{pmatrix}
%   = \frac{\sigma^2}2 I
%   + \frac{\sigma^2 \lvert \rho \rvert}2 
%   \begin{pmatrix}
%     \cos\theta & \sin\theta \\
%     \sin\theta & - \cos\theta
%   \end{pmatrix}
% $$

@@

The expression of Kullback-Leibler divergence of $q(\omega)$ from an arbitrary prior $\pi(\omega)$
necessarily involves the (differential) entropy of $q$:
\begin{equation*}
  KL\bigl(
    q(\omega) \| \pi(\omega)
  \bigr)
    % = \mathbb{E}_{\omega \sim q} \log{q(\omega)} - \mathbb{E}_{\omega \sim q} \log{\pi(\omega)}
    = - \mathbb{H}(q) - \mathbb{E}_{\omega \sim q} \log{\pi(\omega)}
    \,.
\end{equation*}
